### 2025-06-17 basic-statistics.qmd
20250616 - update formulas

- mean (aka "average")
- median
- Hodges-Lehmann statistic (aka pseudomedian)
- mode
- weighted mean
- weighted median

The mean of $x$ (written as: $\bar{x}$) is calculated as in @eq-IQR:

$$
\bar{x} = \frac{\sum x_i}{n} = \frac{x_1 + x_2 + ... + x_n}{n}
$$ {#eq-IQR}

@@

- standard deviation
- variance
- range
- minimum and maximum
- interquartile range (IQR)
- median absolute deviation
- coefficient of variation

@@

where $x_i$ is each data point, $\bar{x}$ is the mean of $x$, and $n$ is the number of data points.

@@

@@

The coefficient of variation (CV) accounts for the fact that variables with larger means also tend to have larger standard deviations; it is calculated as the standard deviation divided by the mean, as in @eq-coefficientOfVariation:

$$
\text{CV} = \frac{SD(x)}{\bar{x}}
$$ {#eq-coefficientOfVariation}

where $SD(x)$ is the standard deviation of $x$.

@@

$$
z = \frac{x - \bar{x}}{\sigma}
$$ {#eq-zScore}

where $x$ is the observed score, $\bar{x}$ is the mean observed score, and $\sigma$ is the standard deviation of the observed scores.

@@

$$
\begin{aligned}
  r &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{the numerator is the covariance; the denominator is the standard deviations} \\
    &= \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{$n-1$ cancels out in numerator and denominator (typical formula for Pearson correlation)} \\
    &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{\sigma_x \cdot \sigma_y} 
  && \quad \text{adding back $n-1$, replacing denominator from line 1 with standard deviation symbols} \\
    &= \frac{\text{Cov}(x, y)}{\sigma_x \cdot \sigma_y} 
  && \quad \text{replacing numerator with covariance notation}
\end{aligned}
$$ {#eq-pearsonCorrelation}

where $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), and $\text{Cov}(x, y)$ is the covariance between $x$ and $y$.

>>>>

- mean (aka "average") (i.e., $M$ or $\mu$)
- median
- Hodges-Lehmann statistic (aka pseudomedian)
- mode
- weighted mean
- weighted median

The mean of $x$ (written as: $\bar{x}$ or $\mu_x$) is calculated as in @eq-mean:

$$
\bar{x} = \frac{\sum x_i}{n} = \frac{x_1 + x_2 + ... + x_n}{n}
$$ {#eq-mean}

@@

- standard deviation ($s$ or $\sigma$)
- variance ($s^2$ or $\sigma^2$)
- range
- minimum and maximum
- interquartile range (IQR)
- median absolute deviation
- coefficient of variation

@@

where $x_i$ is each data point, $\bar{x}$ is the mean of $x$, and $n$ is the number of data points.
The population variance divides by $n$ rather than $n-1$ and is written as $\sigma^2$.

@@

The population standard deviation divides by $n$ rather than $n-1$ and is written as $\sigma$.

@@

The coefficient of variation (CV) accounts for the fact that variables with larger means also tend to have larger standard deviations; it is calculated as the standard deviation divided by the mean, as in @eq-coefficientOfVariation.

@@

$$
z = \frac{x - \bar{x}}{s_x}
$$ {#eq-zScore}

where $x$ is the observed score, $\bar{x}$ is the mean observed score, and $s_x$ is the standard deviation of the observed scores.

@@

$$
\begin{aligned}
  r &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{the numerator is the covariance; the denominator is the standard deviations} \\
    &= \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{$n-1$ cancels out in numerator and denominator (typical formula for Pearson correlation)} \\
    &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{s_x \cdot s_y} 
  && \quad \text{adding back $n-1$, replacing denominator from line 1 with standard deviation symbols} \\
    &= \frac{\text{Cov}(x, y)}{s_x \cdot s_y} 
  && \quad \text{replacing numerator with covariance notation}
\end{aligned}
$$ {#eq-pearsonCorrelation}

where $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), $\text{Cov}(x, y)$ is the covariance between $x$ and $y$, $s_x$ is the standard deviation of $x$, and $s_y$ is the standard deviation of $y$.

### 2025-06-17 correlation.qmd
20250616 - update formulas

$$
\begin{aligned}
  \text{Cov}(x, y) &= \frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})\\
  &= r_{xy} \cdot \sigma_x \cdot \sigma_y
\end{aligned}
$$ {#eq-covariance}

where $n$ is the sample size, $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), $x_i$ is the ith observation on the [predictor variable](#sec-correlationalStudy), $y_i$ is the ith observation of the [outcome variable](#sec-correlationalStudy), $\bar{x}$ is the mean of the [predictor variable](#sec-correlationalStudy), $\bar{y}$ is the mean of the [outcome variable](#sec-correlationalStudy), $r_{xy}$ is the correlation between the [predictor](#sec-correlationalStudy) and [outcome](#sec-correlationalStudy) variables, $\sigma_x$ is the standard deviation of the [predictor variable](#sec-correlationalStudy), and $\sigma_y$ is the standard deviation of the [outcome variable](#sec-correlationalStudy).

>>>>

$$
\begin{aligned}
  \text{Cov}(x, y) &= \frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})\\
  &= r_{xy} \cdot s_x \cdot s_y
\end{aligned}
$$ {#eq-covariance}

where $n$ is the sample size, $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), $x_i$ is the ith observation on the [predictor variable](#sec-correlationalStudy), $y_i$ is the ith observation of the [outcome variable](#sec-correlationalStudy), $\bar{x}$ is the mean of the [predictor variable](#sec-correlationalStudy), $\bar{y}$ is the mean of the [outcome variable](#sec-correlationalStudy), $r_{xy}$ is the correlation between the [predictor](#sec-correlationalStudy) and [outcome](#sec-correlationalStudy) variables, $s_x$ is the standard deviation of the [predictor variable](#sec-correlationalStudy), and $s_y$ is the standard deviation of the [outcome variable](#sec-correlationalStudy).

### 2025-06-17 player-evaluation.qmd
20250616 - update formulas

In terms of consistency, how consistent was the player they from game to game and from season to season?
For instance, we could examine the standard deviations of players' fantasy points across games in a given season.
However, the standard deviation tends to be upwardly biased as the mean increases.
So, we can account for the player's mean fantasy points per game by dividing their game-to-game standard deviation of fantasy points ($\sigma$) by their mean fantasy points across games ($\mu$).
This is known as the coefficient of variation (CV), which is provided in @eq-coefficientOfVariation.

$$
CV = \frac{\sigma}{\mu}
$$ {#eq-coefficientOfVariation}

>>>>

In terms of consistency, how consistent was the player they from game to game and from season to season?
For instance, we could examine the standard deviations of players' fantasy points across games in a given season.
However, the standard deviation tends to be upwardly biased as the mean increases.
So, we can account for the player's mean fantasy points per game by dividing their game-to-game standard deviation of fantasy points ($s_x$ or $\sigma_x$) by their mean fantasy points across games ($\bar{x}$ or $\mu_x$).
This is known as the coefficient of variation (CV), which is provided in @eq-coefficientOfVariation.

$$
\begin{aligned}
  \text{CV} &= \frac{s_x}{\bar{x}}\\
  &= \frac{\sigma_x}{\mu_x}
\end{aligned}
$$ {#eq-coefficientOfVariation}
